Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-08-26 paper.tex
Adding a diagram showcasing how a cache lookup is performed.

\begin{abstract}
Large numerical simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and
even PB ranges. Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
interested in out of the box. In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets
stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. Additionally, we introduce an application-aware 
cache for query results, which has little overhead and improves query performance by X times for queries that hit the cache.

@@

@@

The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
asynchronously. Each node begins evaluation of the query by first performing a cache lookup. If the data for the requested field, time-step and spatial region
are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
parameters the query needs to be evaluated from the raw data. In that case the raw data are read into memory with data along the boundary requested from
adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel. Figure \ref{figCacheLookup} illustrates this process.

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only ~4,300 points (or 0.0004\%
of all points) were above the threshold. For the second set a medium threshold was chosen and ~87,000 points (or 0.0081\% of all points) were above the
threshold. Finally, a low threshold was chosen for the last set with ~900,000 points (or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The execution times were averaged over 10 runs. We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 5\% and within
the margin of error.
\kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
to show how the vorticity values are distributed and to get an understanding for the different threshold levels.}

@@

The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we issued queries
with the same threshold values as in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. The computations for all of 
the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.

We observer a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.3 times when going 
to 4 threads and little speedup is 
observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. This is also exhibited in
the large variance that is seen in the running time of the queries. Additionally, SQL
Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation data atoms along the boundary of
each region will be request by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.5 times faster when compared to running a single
thread.
\kk{Should we have scale-out experiments as well?}

>>>>

\begin{abstract}
Large numerical simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and
even PB ranges. Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
interested in out of the box. In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets
stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. We introduce an application-aware 
cache for query results, which has little overhead and improves query performance by X times for queries that hit the cache.

@@

\begin{figure*}
\centering
\includegraphics[width=6.5in]{Figures/cache_diagram.pdf}
\caption{Cache lookup.}
\label{fig:cache_diagram}
\end{figure*}

@@

The overall execution of threshold queries to the JHTDB is the following. The mediator submits a query to each of the database nodes storing the raw data
asynchronously. Each node begins evaluation of the query by first performing a cache lookup. If the data for the requested field, time-step and spatial region
are available in the cache and if the specified threshold is higher than the one stored in the cache the query can be answered from there. The records are
retrieved from the cache and the ones that have a higher value are returned to the mediator and subsequently back to the user. If the data stored in the
cache have a higher threshold than the one requested the cache needs to be updated. Similarly, if the cache does not have an entry for the specified 
parameters the query needs to be evaluated from the raw data. In that case the raw data are read into memory with data along the boundary requested from
adjacent nodes as needed. The specified field is derived at each location on the grid and the norm or absolute value of the field is compared against
the threshold. The locations where the values are higher than the threshold need to then be stored in the cache. If the cache does not have enough space
for the new records, space is freed up by removing the least recently used data.  Reading from, updating or modifying the cache is done within a
transaction with snapshot isolation level. Snapshot isolation allows us to avoid locking the tables that serve as the cache for each transaction. This
provides for a higher degree of parallelism and avoids any potential deadlocks from queries running in parallel. Figure \ref{fig:cache_diagram} illustrates 
this process.

@@

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
The execution times were averaged over 10 runs and cache entries for the particular time-step were dropped before each run. 
We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is small, less than 5\% and within
the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels
\kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
to show how the vorticity values are distributed and to get an understanding for the different threshold levels.}

@@

The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. The computations for all of 
the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.

We observer a two times speedup when going from a single thread per node to two threads per node. The speedup diminishes to 1.3 times when going 
to 4 threads and little speedup is 
observed with 8 threads per node. This is due to several factors. Each node has 2 CPUs and a total of 8 cores, which need to run operating system and other SQL Server
processes as well. The experiments were run on the live production database nodes, which were also servicing other user queries. This is also exhibited in
the large variance that is seen in the running time of the queries. Additionally, SQL
Server already utilizes parallelism to perform most of the I/O. Finally, due to the data-parallel approach to the evaluation of these queries the data atoms along
the boundary of each region will be requested by multiple threads. Nevertheless, running with 4 threads per node is nearly 2.5 times faster when compared to
running a single thread.
\kk{Should we have scale-out experiments as well?}

Fig. \ref{fig:io_comparison} compares the running time of the queries with a medium threshold and executed with varying number of threads per node with
the time taken to perform the I/O only. The I/O time is about half of the total running time for these queries. As expected weaker than linear scale-up is 
observed for the I/O when additional threads are utilized. SQL Server already makes use of parallelism internally and the data have to be retrieve from the
same set of disks. It is worth noting that the total running time for the queries evaluated with 4 or 8 threads is about the same as the time it takes to perform the
I/O only with a single thread.
